---
title: "R Notebook"
output: html_notebook
---

```{r}

library(tidyverse)
library(nycflights13)
library(geosphere)
library(stringr)
library(lubridate)

```



This chunk creates intervals for the flights to fall under
```{r}

intervals <- tibble(
  min_interval = seq(00, 2359, 100),
  max_interval = seq(59, 2359, 100)
)

intervals$max_interval[intervals$max_interval == 2359] = 2400

intervals <- intervals %>% 
  mutate(min_interval = str_pad(intervals$min_interval, 4, "left", 0),
         max_interval = str_pad(intervals$max_interval, 4, "left", 0)) %>% 
  
  mutate(combined = paste0(min_interval,"-",max_interval))

```



This massages the flights df into something usable by my app with minimal work.

```{r}

## from_timing_point = origin
## to_timing_point = dest
## route = carrier

new_flights <- flights %>% 
  mutate(time_hour = ymd_hms(time_hour)) %>% 
  mutate(service_week = wday(time_hour)) %>% 
  mutate(service_type_code = ifelse(service_week == 1, "SUN", ifelse(service_week == 7, "SAT", "MF"))) 

standardized_flights <- inner_join(new_flights, select(airports, faa, tz), by = c('origin' = 'faa')) %>% 
  inner_join(select(airports, faa, tz), by = c('dest' = 'faa')) %>% 
  mutate(from_tz = tz.x,
         to_tz = tz.y)


new_std_flights <- standardized_flights %>%
  mutate(sched_dep_time_sec = (sched_dep_time %/% 100)*60*60 + (sched_dep_time %% 100)*60 + from_tz*60*60 ) %>% 
  mutate(sched_arr_time_sec = (sched_arr_time %/% 100)*60*60 + (sched_arr_time %% 100)*60 + to_tz*60*60 ) %>% 
  mutate(dep_time_sec = (dep_time %/% 100)*60*60 + (dep_time %% 100)*60 + from_tz*60*60) %>% 
  mutate(arr_time_sec = (arr_time %/% 100)*60*60 + (arr_time %% 100)*60 + to_tz*60*60) %>% 
  mutate(sch_rt = abs(sched_dep_time_sec - sched_arr_time_sec)) %>% 
  mutate(act_rt = abs(dep_time_sec - arr_time_sec))
  

```

```{r}
getSeason <- function(input.date){
  numeric.date <- 100*month(input.date)+day(input.date)
  ## input Seasons upper limits in the form MMDD in the "break =" option:
  cuts <- base::cut(numeric.date, breaks = c(0,319,0620,0921,1220,1231)) 
  # rename the resulting groups (could've been done within cut(...levels=) if "Winter" wasn't double
  levels(cuts) <- c("Winter","Spring","Summer","Fall","Winter")
  return(cuts)
}

# Function courtesy of https://stackoverflow.com/questions/36502140/determine-season-from-date-using-lubridate-in-r#36518085


season_dates <- tibble(
  season = c("Winter","Spring","Summer","Fall"),
  min_date = ymd_hms(c("2013-12-20 00:00:00", "2013-03-19 00:00:00", "2013-06-20 00:00:00", "2013-09-21 00:00:00")),
  max_date = ymd_hms(c("2013-03-18 12:59:59", "2013-06-19 12:59:59", "2013-09-20 12:59:59", "2013-12-19 12:59:59")) )

```


```{r}

new_std_flights <- new_std_flights %>% 
  mutate(season = getSeason(time_hour))

```



Group into intervals by hour
```{r}
suppressWarnings(computed_intervals <- unlist(map(.x = new_std_flights$dep_time, .f = function(x) intervals$combined[min(which(x < as.numeric(intervals$max_interval)))] )))

new_std_flights$interval <- computed_intervals
```


This creates a dataframe of all existing combinations of from and to airports and get's both lat and lons.
```{r}

unique_paths <- unique(flights[,c('origin','dest')])

pre_direc <- inner_join(unique_paths, select(airports, faa, lat, lon), by = c('origin' =  'faa')) %>% 
  inner_join( select(airports, faa, lat, lon), by = c('dest' =  'faa')) 

```

Now we can figure out what the general direction of the flight is going. We first find the initial angle of the flight and then attribute to them cardinal directions.

```{r}


pre_direc$angle <- bearingRhumb(p1 = as.matrix(tibble(pre_direc$lon.x, pre_direc$lat.x)), 
             p2 = as.matrix(tibble(pre_direc$lon.y, pre_direc$lat.y)))


dir_sequence <- seq(-22.5,360, 45)


coversion_table <- tibble(start = dir_sequence, 
       end = lead(dir_sequence, default = 360),
       direction_name = c("E","NE","N","NW","W","SW","S","SE","E"))

suppressWarnings(computed_direction <- unlist(map(.x = pre_direc$angle, .f = function(x) coversion_table$direction_name[min(which(x < as.numeric(coversion_table$end)))] )))

pre_direc$direction_name <- computed_direction
```

Now that we know what direction a flight will be going we want to attach that to the flights people might take. But we don't really care to attributed a direction to each entry, just to each unique one .

SELECT distinct line_no, direction_name, service_type_code, from_timing_point, to_timing_point,

```{r}

flight_options <- new_std_flights %>% 
  group_by(origin, dest, service_type_code, carrier, interval) %>% 
  summarise_each(funs(first)) %>% 
  ungroup() %>% 
  left_join(pre_direc) %>% 
  select(origin, dest, service_type_code, carrier, interval, direction_name)


```

save this data for use in the app in the future. In the future we want to add more years worth of data.
```{r}
save(flight_options, file = "../data/flight_options.rda")
```


to create historical predictions
```{r}

flight_history <- new_std_flights %>% 
  group_by(origin, dest, service_type_code, carrier, interval, season) %>% 
  summarise(sch_rt = mean(sch_rt),
            mean_act_rt = mean(act_rt, na.rm = TRUE),
            median_act_rt = median(act_rt, na.rm = TRUE)
            ) %>% 
  ungroup() %>% 
  left_join(pre_direc) %>% 
  select(origin, dest, service_type_code, carrier, interval, season, direction_name, sch_rt, mean_act_rt, median_act_rt)

flight_history <- left_join(flight_history, season_dates, by = 'season')

```


```{r}

to_minutes <- 60


fh <- flight_history %>% 
  filter(origin == "EWR",
         dest == "ALB",
         service_type_code == "MF",
         carrier == "EV",
         interval == "0700-0759",
         direction_name == "E"
         ) %>% 
  gather("minormax", "date", min_date, max_date)

save(fh, file = "../data/fh.rda")
      

```



```{r}
ggplot(data = fh) +
          geom_line(aes(x = date, y = sch_rt/to_minutes, color = I("#3019ff") ), linetype = "dashed") +
  geom_line(aes(x = date, y = mean_act_rt/to_minutes, color = I("#3019ff") ), linetype = "solid") 

```






```{r}
flights_use <- new_std_flights %>% 
  select(origin,dest,service_type_code, sch_rt, act_rt, carrier, interval)
```











to grab more data use hadley's code here and update year: https://github.com/hadley/nycflights13/blob/master/data-raw/flights.R






